---
title: "Covid 19 Data Analysis"
output:
  pdf_document: default
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Johns Hopkins University COVID-19 Data Repository 

This report details the analysis of COVID-19 Data from daily reports that Johns Hopkins University maintained until March 10, 2023. 

First we will start with the data loading from the Johns Hopkins University Covid 19 Data on Github, we will be downloading the raw file to pull into R markdown. Then we will clean the data. Next we will do some visual analysis and modeling. 

*On March 10, 2023, the Johns Hopkins Coronavirus Resource Center ceased its collecting and reporting of global COVID-19 data. The current data from this source is archived data. We are analyzing the time series data from Global and US data regarding cases and deaths.*


- **Non-Base Packages and Versions Used for Analysis:**
   - `forecast`: 8.21.1
   - `RColorBrewer`: 1.1-3
   - `knitr`: 1.44
   - `lubridate`: 1.9.3
   - `forcats`: 1.0.0
   - `stringr`: 1.5.0
   - `dplyr`: 1.1.3
   - `purrr`: 1.0.2
   - `readr`: 2.1.4
   - `tidyr`: 1.3.0
   - `tibble`: 3.2.1
   - `ggplot2`: 3.4.4
   - `tidyverse`: 2.0.0

### Package loading
```{r package_loading, message = FALSE}
#Loading all packages
library(tidyverse) 
library(dplyr) 
library(lubridate) 
library(knitr) 
library(ggplot2) 
library(RColorBrewer) 
library(forecast)

```

# Data Loading

We will use the main link to string concatenate each file of data, allowing a more systematic approach to loading all of the CSV files. 
```{r load_data,  message = FALSE}

#Main url link
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

#Links to each data source
file_names <- c("time_series_covid19_confirmed_US.csv",  
                "time_series_covid19_confirmed_global.csv", 
                "time_series_covid19_deaths_US.csv",  
                "time_series_covid19_deaths_global.csv")

#Concat each link 
urls <- str_c(url_in, file_names)

global_cases <- read_csv(urls[2])
global_deaths <- read_csv(urls[4])
us_cases <- read_csv(urls[1])
us_deaths <- read_csv(urls[3])

#UID for country populations lookup
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"


#Reading and cleaning part of uid
uid <- read.csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2)) %>%
  mutate(Province_State = trimws(toupper(Province_State)),
         Country_Region = trimws(toupper(Country_Region)))
```

# Data Cleaning and Tidying

We are going to tidy the data for each of the links. We are going to make sure each variable has its own column and make sure all of the data types are correct. Some columns are also going to be dropped due to being unneeded such as Latitude and Longitude. Dates are pivoted into rows. 

## Global Data Cleaning
```{r tidy_global_data, message= FALSE, warning = FALSE}

#Clean Global Case Data 
global_cases <- global_cases %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long), 
               names_to = "date", 
               values_to = "cases") %>%
  rename(Province_State = `Province/State`,
         Country_Region = `Country/Region`) %>%
  mutate(Province_State = ifelse(is.na(Province_State), Province_State, trimws(toupper(Province_State))),
         Country_Region = trimws(toupper(Country_Region))) %>%
  select(-c(Lat, Long))

#Clean Global Death Data
global_deaths <- global_deaths %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long), 
               names_to = "date", 
               values_to = "deaths") %>%
  rename(Province_State = `Province/State`,
         Country_Region = `Country/Region`) %>%
  mutate(Province_State = ifelse(is.na(Province_State), Province_State, trimws(toupper(Province_State))),
         Country_Region = trimws(toupper(Country_Region))) %>%
  select(-c(Lat, Long))

#Combine Global cases and deaths 
global <- global_cases %>% 
  full_join(global_deaths) %>%
  mutate(date = mdy(date))

#Filter for start of Data 
global <- global %>% filter(cases > 0)

global <- global %>%  
  unite("Combined_Key", 
        c(Province_State, Country_Region),
        sep = ", ", 
        na.rm = TRUE, 
        remove = FALSE)

#Perform the join focusing on the Country_Region
global <- global %>%
  left_join(uid, by = "Country_Region") %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State.x, Country_Region, date, cases, deaths, Population, Combined_Key)

#Rename the columns for clarity
names(global)[names(global) == "Province_State.x"] <- "Province_State"


head(global)

```

## US Data Cleaning
```{r tidy_US_data, message= FALSE}
#Clean US Case Data 
us_cases <- us_cases %>%
  pivot_longer(cols = -(UID:Combined_Key), 
               names_to = "date", 
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))



#Clean US Death Data 
us_deaths <- us_deaths %>%
  pivot_longer(cols = -(UID:Population), 
               names_to = "date", 
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

#Make a full US dataframe
US <- us_cases %>%
  full_join(us_deaths)

head(US)
```


# Visualizing Data 

Next we will explore the tidy data from the previous sections to make an analysis of the data with visualizations and models. First we will be plotting Global Cases and Deaths Over time. Next we plot the timeline of case reporting in India and China due to their similarity in population. 

## Global Data Aggregation Visualizations

```{r global_plotting, message= FALSE}

#Data aggregation
aggregated_data <- global %>% 
  group_by(date) %>%
  summarise(total_cases = sum(cases, na.rm = TRUE), 
            total_deaths = sum(deaths, na.rm = TRUE))

#Plot for Cases
p1 <- ggplot(aggregated_data, aes(x = date, y = total_cases)) +
  geom_line(color = "blue") +
  labs(title = "Global COVID-19 Cases Over Time",
       x = "Date",
       y = "Total Cases") +
  theme_minimal()

#Plot for Deaths
p2 <- ggplot(aggregated_data, aes(x = date, y = total_deaths)) +
  geom_line(color = "red") +
  labs(title = "Global COVID-19 Deaths Over Time",
       x = "Date",
       y = "Total Deaths") +
  theme_minimal()

#Display the plots of Global Cases and Deaths
gridExtra::grid.arrange(p1, p2, ncol = 1)
```

**Figure 1,2: For the Global Cases, the number of cases continue to rise. There is a notable jump in cases in early 2022 that we are going to explore further. The Global Deaths chart does not reflect the same jump in early 2022** 

## India and China Data Aggregation Visualization

```{r country_plotting, message= FALSE}
#Filtering and aggregating the data for India
india_data <- global %>% 
  filter(Country_Region == "INDIA") %>%
  group_by(date) %>%
  summarise(total_cases = sum(cases, na.rm = TRUE), 
            total_deaths = sum(deaths, na.rm = TRUE))

#Filtering and aggregating the data for China
china_data <- global %>% 
  filter(Country_Region == "CHINA") %>%
  group_by(date) %>%
  summarise(total_cases = sum(cases, na.rm = TRUE), 
            total_deaths = sum(deaths, na.rm = TRUE))

combined_data <- bind_rows(
  india_data %>% mutate(Country = "India"),
  china_data %>% mutate(Country = "China")
)

p3 <- ggplot(combined_data, aes(x = date, y = total_cases, color = Country)) +
  geom_line() +
  facet_wrap(~Country, scales = "free_y") +
  labs(title = "Comparison of COVID-19 Cases: India vs. China",
       x = "Date",
       y = "Total Cases") +
  theme_minimal()


#Display the plot of China and India cases 
p3

```

**Figure 3: In this graph we can see the reporting of COVID-19 Cases of the two most populous countries, China and India. China is the original country to have COVID-19 and had a very robust testing system early 2020, but did not share numbers until 2022. This can be seen on the chart on the left above, the cases were flat until early 2022, which explains why the global cases saw a peak in Figure 1 around early 2022. This could cause some bias in any Global modeling that could have occurred prior to 2022.**


# Modeling Data 

In this section, a model was created to forecast Global Cases 60 days past date that reporting was concluded on March 10, 2023. 

## Forecasting Data 60 Days past the reporting endtime 
```{r models, message = FALSE}

#Aggregating global data
global_aggregated <- global %>%
  group_by(date) %>%
  summarise(total_cases = sum(cases, na.rm = TRUE))

#Converting to a time series object
cases_ts <- ts(global_aggregated$total_cases, frequency = 7)

#Fitting an ARIMA model
arima_model <- auto.arima(cases_ts)

#Forecasting the next 60 days
forecast_values <- forecast(arima_model, h = 60)

#Extracting the last year of actual data
last_year_data <- tail(global_aggregated, 365)

#Creating a sequence of dates for the forecast
forecast_dates <- seq(from = as.Date("2023-03-11"), by = "days", length.out = 60)

#Combining the actual and forecasted data
combined_data <- data.frame(
  date = c(last_year_data$date, forecast_dates),
  cases = c(last_year_data$total_cases, forecast_values$mean)
)

#Creating the plot
p <- ggplot() +
  geom_line(data = combined_data[1:nrow(last_year_data), ], aes(x = date, y = cases), color = "blue") +
  geom_line(data = combined_data[(nrow(last_year_data)+1):nrow(combined_data), ], aes(x = date, y = cases), color = "red") +
  labs(title = "60-day Forecast After Data Collection Ended",
       x = "Date",
       y = "Total Cases") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


print(p)

```
**Figure 4: In this figure, the red portion of the line represents the forecasted cases 60 days past the final reporting date March 10, 2023. We can still see in this that cases are expected to rise over this period.**

# Conclusion

This report looks at COVID-19 data up until March 10, 2023, from Johns Hopkins University. Here is what we found:

1. **Increasing Cases:** Globally, COVID-19 cases are still going up. There was a big jump in reported cases in early 2022, but we did not see the same jump in the number of deaths (Figures 1 and 2).

2. **China and India’s Cases:** We compared COVID-19 cases in China and India, the two biggest countries by population (Figure 3). China, where the virus first started, did not report many cases until 2022. This sudden increase in reports in 2022 is why we see a big jump in global cases around the same time. This means that any world predictions made before 2022 might not be accurate.

3. **What is Next:** We predicted the number of cases for the next 60 days after March 10, 2023 (Figure 4). It looks like the number of cases keep going up during this time.

Our study shows that at that time COVID-19 was still spreading, and the numbers were increasing. Changes in how cases were reported, especially in China, make it a bit tricky to understand the true global picture. We should keep in mind that sudden changes in reported cases can bias our models and understanding of the virus's spread.


